Exploring the Utility of Machine Learning Across Varied Data Formats

### 2024-05-23 homework2/paper/paper.tex
finish second chapter

Using the \textit{analysys\_attributes.py} script, we can identify the 
Continuous Numeric Attributes and Discrete Nominal Attributes in a specific 
dataset. The script will output statistics that can be showed in Tables 
\ref{tab:continuous_numeric_attributes_salary} and 
\ref{tab:continuous_numeric_attributes_stroke} for numeric attributes and
Table \ref{tab:discrete_nominal_attributes_salary} and 
\ref{tab:discrete_nominal_attributes_stroke} for nominal attributes.

@@

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{1cm}||>{\centering\arraybackslash}m{2.25cm}|>{\centering\arraybackslash}m{2.25cm}|>{\centering\arraybackslash}m{2.25cm}|>{\centering\arraybackslash}m{2.25cm}|>{\centering\arraybackslash}m{2.25cm}|}
        \hline
        \multicolumn{6}{|c|}{List of all Continuous Numeric Attributes in the Stroke Prediction dataset} \\
        \hline
        & mean\_blood\_ sugar\_level & cardiovascular\_ issues & body\_mass\_ indicator & high\_ blood\_pressure & years\_old \\
        \hline\hline
        count & 5110.000000 & 5110.000000 & 4909.000000 & 5110.000000 & 5110.000000 \\
        \hline
        mean & 106.147677 & 0.054012 & 28.893237 & 0.097456 & 46.568665 \\
        \hline
        std & 45.283560 & 0.226063 & 7.854067 & 0.296607 & 26.593912 \\
        \hline
        min & 55.120000 & 0.000000 & 10.300000 & 0.000000 & 0.080000 \\
        \hline
        25\% & 77.245000 & 0.000000 & 23.500000 & 0.000000 & 26.000000 \\
        \hline
        50\% & 91.885000 & 0.000000 & 28.100000 & 0.000000 & 47.000000 \\
        \hline
        75\% & 114.090000 & 0.000000 & 33.100000 & 0.000000 & 63.750000 \\
        \hline
        max & 271.740000 & 1.000000 & 97.600000 & 1.000000 & 134.000000 \\
        \hline
    \end{longtable}
    \begin{longtable}{ |>{\centering\arraybackslash}m{1cm}||>{\centering\arraybackslash}m{2.5cm}|>{\centering\arraybackslash}m{2.5cm}|>{\centering\arraybackslash}m{2.5cm}|>{\centering\arraybackslash}m{2.5cm}|}
        \hline
        \multicolumn{5}{|c|}{List of all Continuous Numeric Attributes in the Stroke Prediction dataset} \\
        \hline
        & chaotic\_sleep & analysis\_ results & biological\_ age\_index & cerebrovascular\_ accident \\
        \hline\hline
        count & 5110.000000 & 4599.000000 & 5110.000000 & 5110.000000 \\
        \hline
        mean & 0.054012 & 323.523446 & 134.784256 & 0.048728 \\
        \hline
        std & 0.226063 & 101.577442 & 50.399352 & 0.215320 \\
        \hline
        min & 0.000000 & 104.829714 & -15.109456 & 0.000000 \\
        \hline
        25\% & 0.000000 & 254.646209 & 96.710581 & 0.000000 \\
        \hline
        \hline
        50\% & 0.000000 & 301.031628 & 136.374631 & 0.000000 \\
        \hline
        \hline
        75\% & 0.000000 & 362.822769 & 172.507322 & 0.000000 \\
        \hline
        max & 1.000000 & 756.807975 & 266.986321 & 1.000000 \\
        \hline
        \caption{Continuous Numeric Attributes in Stroke Prediction Dataset}
        \label{tab:continuous_numeric_attributes_stroke} \\
   \end{longtable}
\end{center}

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{5cm}|||>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|}
        \hline
        \multicolumn{3}{|c|}{List of all Discrete Nominal Attributes in the Salary Prediction dataset} \\
        \hline
        & Non-missing count & Unique values count \\
        \hline\hline
        \hline
        country & 9999 & 41 \\
        \hline
        job & 9999 & 14 \\
        \hline
        work\_type & 9999 & 9 \\
        \hline
        partner & 9999 & 7 \\
        \hline
        edu & 9999 & 16 \\
        \hline
        gender & 9199 & 2 \\
        \hline
        race & 9999 & 5 \\
        \hline
        gtype & 9999 & 2 \\
        \hline
        money & 9999 & 2 \\
        \hline
        \caption{Discrete Nominal Attributes in Salary Prediction Dataset}
        \label{tab:discrete_nominal_attributes_salary} \\
    \end{longtable}
\end{center}

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{5cm}|||>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|}
        \hline
        \multicolumn{3}{|c|}{List of all Discrete Nominal Attributes in the Stroke Prediction dataset} \\
        \hline
        & Non-missing count & Unique values count \\
        \hline\hline
        job\_category & 5110 & 5 \\
        \hline
        sex & 5110 & 2 \\
        \hline
        tobacco\_usage & 5110 & 4 \\
        \hline
        married & 4599 &2 \\
        \hline
        living\_area & 5110 & 2 \\
        \hline

@@

\subsection{Investigation of Class Distribution}
TODO

\subsection{Analysis of Feature Correlations}
TODO

@@

% ---- Bibliography ----
\begin{thebibliography}{8}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/A*_search_algorithm#Complexity}{A* search algorithm - Wikipedia}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Hill_climbing}{Hill Climbing - Wikipedia}
    \bibitem{}
    \href{https://curs.upb.ro/2023/course/view.php?id=13749}{Moodle - Artifical Intelligence Course}
    \bibitem{}
    \href{https://www.researchgate.net/profile/Kian-Pokorny/publication/262277120_Multiple_constraint_satisfaction_problems_using_the_A-star_A_search_algorithm_classroom_scheduling_with_preferences/links/5e57f530299bf1bdb8408b03/Multiple-constraint-satisfaction-problems-using-the-A-star-A-search-algorithm-classroom-scheduling-with-preferences.pdf}{Multiple constraint satisfaction problems using the A-star search algorithm}
    \bibitem{}
    \href{https://d1wqtxts1xzle7.cloudfront.net/68245993/14443-libre.pdf?1626958962=&response-content-disposition=inline%3B+filename%3DCourses_timetabling_based_on_hill_climbi.pdf&Expires=1714507660&Signature=JDd-gaDdE4E7s5P66Op535FyEuysvPoA5o4fcwak4NWT1sQNtVh50ooYa9eSrHu~aHR-Ulabenz910uDjozKpgHivCFkhipTT5bJJL9MensZTSib8TBVzGtBU9lycqCVwilt21Udi0DMgfhERDSgCcR4vO8hmlFlxOLg1exD1fNLE8kxt4S0H2Qfb3nQEHu3kLsicuJNk8wfF~eWpEDUlWsjW9J9IpxuLWP0OICQ7ZReWtx37dHnPN2PTVmFs815NMNi-jyLV0gWWBTAOL7Hn7A5VSg-l~s2e8RAoCjPAYQDRnnznIzOElRy2nTsgkW3mZDzlsntJ3lFur2KquwqNg__&Key-Pair-Id=APKAJLOHF5GGSLRBV4ZA}{Courses timetabling based on hill climbing algorithm}

>>>>

Using the \textit{analysis\_attributes.py} script, we can identify the 
Continuous Numeric Attributes and Discrete Nominal Attributes in a specific 
dataset. The script will output statistics that can be showed in Tables 
\ref{tab:continuous_numeric_attributes_salary} and 
\ref{tab:continuous_numeric_attributes_stroke} for numeric attributes and
Table \ref{tab:discrete_nominal_attributes_salary} and 
\ref{tab:discrete_nominal_attributes_stroke} for nominal attributes.

@@

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{1cm}||>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|}
        \hline
        \multicolumn{5}{|c|}{List of all Continuous Numeric Attributes in the Stroke Prediction dataset} \\
        \hline
        & mean\_blood\_ sugar\_level & cardiovascular\_ issues & body\_mass\_ indicator & high\_ blood\_pressure \\
        \hline\hline
        count & 5110.000000 & 5110.000000 & 4909.000000 & 5110.000000 \\
        \hline
        mean & 106.147677 & 0.054012 & 28.893237 & 0.097456 \\
        \hline
        std & 45.283560 & 0.226063 & 7.854067 & 0.296607 \\
        \hline
        min & 55.120000 & 0.000000 & 10.300000 & 0.000000 \\
        \hline
        25\% & 77.245000 & 0.000000 & 23.500000 & 0.000000 \\
        \hline
        50\% & 91.885000 & 0.000000 & 28.100000 & 0.000000 \\
        \hline
        75\% & 114.090000 & 0.000000 & 33.100000 & 0.000000 \\
        \hline
        max & 271.740000 & 1.000000 & 97.600000 & 1.000000 \\
        \hline
    \end{longtable}
    \begin{longtable}{ |>{\centering\arraybackslash}m{1cm}||>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|}
        \hline
        \multicolumn{5}{|c|}{List of all Continuous Numeric Attributes in the Stroke Prediction dataset} \\
        \hline
        & years\_old & chaotic\_sleep & analysis\_results & biological\_age\_index \\
        \hline\hline
        count & 5110.000000 & 5110.000000 & 4599.000000 & 5110.000000 \\
        \hline
        mean & 46.568665 & 0.054012 & 323.523446 & 134.784256 \\
        \hline
        std & 26.593912 & 0.226063 & 101.577442 & 50.399352 \\
        \hline
        min & 0.080000 & 0.000000 & 104.829714 & -15.109456 \\
        \hline
        25\% & 26.000000 & 0.000000 & 254.646209 & 96.710581 \\
        \hline
        50\% & 47.000000 & 0.000000 & 301.031628 & 136.374631 \\
        \hline
        75\% & 63.750000 & 0.000000 & 362.822769 & 172.507322 \\
        \hline
        max & 134.000000 & 1.000000 & 756.807975 & 266.986321 \\
        \hline
        \caption{Continuous Numeric Attributes in Stroke Prediction Dataset}
        \label{tab:continuous_numeric_attributes_stroke} \\
   \end{longtable}
\end{center}

An initial inspection of the data reveals that there are missing attributes in 
both the Salary Prediction and Stroke Prediction datasets. The Salary Prediction 
dataset is missing the \textit{'hpw'} attribute, while the Stroke Prediction dataset is
missing two attributes: \textit{'body\_mass\_indicator'} and 
\textit{'analysis\_results'}.

To better understand the distribution of the continuous numeric attributes 
within the datasets, boxplots have been generated for each attribute. These 
visualizations are located in the \textit{'plots'} folder at the root of the 
project directory. The name of each boxplot starts with \textit{'box\_plot\_'}.

Boxplots are a standardized method for visually representing the distribution of
data.  They provide insights into several key characteristics of the data, 
including the median, quartiles, and outliers.

In the Figure \ref{fig:boxplot_example_salary} we can see a boxplot for the 
\textit{years} attribute in the Salary Prediction dataset. The box in the middle
of the plot contains the middle 50\% of the data, and the line in the middle
represents the median. The whiskers extend to the minimum and maximum values
within 1.5 times the interquartile range (the difference between the first and
third quartiles). Points outside this range are considered outliers.

Also, in Figure \ref{fig:boxplot_example_stroke} we can see a boxplot for the
\textit{body\_mass\_indicator} attribute in the Stroke Prediction dataset.
As described above, the boxplot provides a visual representation of the data's
distribution, highlighting key statistical measures such as the median, quartiles,
and potential outliers. These information is also presented in the tables
\ref{tab:continuous_numeric_attributes_salary} and
\ref{tab:continuous_numeric_attributes_stroke}. One of the main insights that
can be derived from the boxplot is the presence of outliers, which are data points
that lie significantly outside the range of the rest of the data. Outliers can
have a significant impact on the performance of machine learning models, and
identifying and handling them appropriately is an essential step in the data
preprocessing process.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/box_plot_years_SalaryPrediction_full.png}
    \caption{Boxplot for the \textit{years} attribute in the Salary Prediction dataset}
    \label{fig:boxplot_example_salary}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/box_plot_body_mass_indicator_AVC_full.png}
    \caption{Boxplot for the \textit{body\_mass\_indicator} attribute in the Stroke Prediction dataset}
    \label{fig:boxplot_example_stroke}
\end{figure}

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{5cm}||>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|}
        \hline
        \multicolumn{3}{|c|}{List of all Discrete Nominal Attributes in the Salary Prediction dataset} \\
        \hline
        & Non-missing count & Unique values count \\
        \hline\hline
        relation & 9999 & 6 \\
        \hline
        country & 9999 & 41 \\
        \hline
        job & 9999 & 14 \\
        \hline
        work\_type & 9999 & 9 \\
        \hline
        partner & 9999 & 7 \\
        \hline
        edu & 9999 & 16 \\
        \hline
        gender & 9199 & 2 \\
        \hline
        race & 9999 & 5 \\
        \hline
        gtype & 9999 & 2 \\
        \hline
        \caption{Discrete Nominal Attributes in Salary Prediction Dataset}
        \label{tab:discrete_nominal_attributes_salary} \\
    \end{longtable}
\end{center}

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{5cm}||>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|}
        \hline
        \multicolumn{3}{|c|}{List of all Discrete Nominal Attributes in the Stroke Prediction dataset} \\
        \hline
        & Non-missing count & Unique values count \\
        \hline\hline
        job\_category & 5110 & 5 \\
        \hline
        sex & 5110 & 2 \\
        \hline
        tobacco\_usage & 5110 & 4 \\
        \hline
        married & 4599 &2 \\
        \hline
        living\_area & 5110 & 2 \\
        \hline

@@

From the Discret Nominal Attributes tables we can see that each dataset contains
only one attribute with missing values. The Salary Prediction dataset is missing
the \textit{gender} attribute, while the Stroke Prediction dataset is missing the
\textit{married} attribute. Also, the number of unique values for each attribute
descibes the diversity of the data. For example, the \textit{country} attribute
in the Salary Prediction dataset has 41 unique values, indicating that the data
contains information from 41 different countries.

In the historigrams for the discrete nominal attributes, we can see the
distribution of the unique values for each attribute. These visualizations can
provide insights into the frequency of each category within the dataset, which
can be useful for understanding the data's composition and identifying potential
imbalances or biases. The histograms for the discrete nominal attributes are
located in the \textit{'plots'} folder at the root of the project directory.
The name of each histogram starts with \textit{'histogram\_'}.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/histogram_work_type_SalaryPrediction_full.png}
    \caption{Histogram for the \textit{work\_type} attribute in the Salary Prediction dataset}
    \label{fig:histogram_example_salary}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/histogram_tobacco_usage_AVC_full.png}
    \caption{Histogram for the \textit{tobacco\_usage} attribute in the Stroke Prediction dataset}
    \label{fig:histogram_example_stroke}
\end{figure}

In Figure \ref{fig:histogram_example_salary} we can see a histogram of the
\textit{work\_type} attribute in the Salary Prediction dataset. The dominance of
the 'Priv' category indicates a severe class imbalance. For classification tasks,
the model might predict Priv most of the time since it's the majority class, 
leading to a high overall accuracy but poor precision, recall, and F1 scores 
for minority classes.

Also, in Figure \ref{fig:histogram_example_stroke} we can see a histogram of the
\textit{tobacco\_usage} attribute in the Stroke Prediction dataset. The histogram 
shows that the majority of individuals are non-smokers, with a significant portion 
having undefined tobacco usage status. This imbalance and the presence of missing 
data need to be addressed appropriately.

\subsection{Investigation of Class Distribution}
In machine learning, it is common practice to split a dataset into two distinct
subsets: a training set and a test set. This division is crucial for ensuring 
 robustness and generalizability of the models developed using the data.

\begin{itemize}
    \item \textbf{Training Set:} The primary purpose of the training set is to 
    train the machine 
    learning model. The model learns from patterns and relationships within the data 
    to develop a predictive capability.
    \item \textbf{Test Set:} The test set, unseen by the model during training, 
    serves to evaluate the model's generalizability. By applying the trained 
    model to the test set, we can assess its performance on new, unseen data. 
    This helps prevent overfitting, where the model performs well on the 
    training data but fails to generalize to real-world scenarios.
\end{itemize}

Looking at how data is distributed is key. Imbalanced data, where some classes 
have far more examples than others, throws off classification tasks:
high accuracy can hide poor performance on rare classes; models struggle to 
learn patterns from underrepresented classes; inaccurate predictions, especially 
for the minority class.

By checking the distribution, we can address imbalance:
\begin{itemize}
    \item Balance the data: Oversample rare examples or undersample common ones.
    \item Cost-sensitive learning: Penalize the model more for mistakes on rare classes.
    \item Better metrics: Use precision, recall, and F1-score to get a clearer picture.
\end{itemize}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/distribution_edu_Salary_Prediction.png}
    \caption{Distribution of the \textit{edu} attribute in the Salary Prediction dataset}
    \label{fig:distribution_example_salary}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/distribution_tobacco_usage_Stroke_Prediction.png}
    \caption{Distribution of the \textit{tobacco\_usage} attribute in the Stroke Prediction dataset}
    \label{fig:distribution_example_stroke}
\end{figure}

In the \textit{'plots'} folder at the root of the project directory we can find
the distribution of each class in the datasets. The name of each plot starts
with \textit{'distribution\_'}. Here is an example of the distribution of the
\textit{edu} attribute in the Salary Prediction dataset in Figure
\ref{fig:distribution_example_salary}. Also, the distribution of the
\textit{tobacco\_usage} attribute in the Stroke Prediction dataset is shown in
Figure \ref{fig:distribution_example_stroke}.

\subsection{Analysis of Feature Correlations}
Feature correlation analysis is a critical step in understanding the relationships
between different attributes in a dataset. By examining how attributes are related
to each other, we can identify patterns, dependencies, and redundancies that can
inform feature selection, model building, and interpretation.

Correlation analysis typically involves calculating correlation coefficients
between pairs of attributes. The correlation coefficient quantifies the strength
and direction of the linear relationship between two variables. A correlation
coefficient close to 1 indicates a strong positive relationship, while a value
close to -1 indicates a strong negative relationship. A correlation coefficient
near 0 suggests no linear relationship between the variables.

In the \textit{'correlation\_analysis.py'} script, we calculate the correlation
coefficients between all pairs of continuous numeric attributes in the datasets,
generating a correlation matrix for each dataset. Moreover, we calculate the 
Cramér's V coefficient for all pairs of discrete nominal attributes in the datasets,
generating a Cramér's V matrix for each dataset to measure the association between
categorical variables. In Figures \ref{fig:correlation_matrix_example_salary} and
\ref{fig:correlation_matrix_example_stroke} we can see the correlation matrix for
the Salary Prediction and Stroke Prediction datasets, respectively, for the continuous
numeric attributes. In Figures \ref{fig:cramer_v_matrix_example_salary} and
\ref{fig:cramer_v_matrix_example_stroke} we can see the Cramér's V matrix for
the discrete nominal attributes.

The correlation matrix and Cramér's V matrix provide valuable insights into the
relationships between attributes in the datasets.By examining these matrices, we
can see that Figure \ref{fig:correlation_matrix_example_salary} the \textit{prod}
attribute is highly correlated with the \textit{gain} attribute, while the
\textit{years} attribute is negatively correlated with the \textit{fnl} attribute.

In Figure \ref{fig:correlation_matrix_example_stroke} we can see that the
\textit{cardiovascular\_issues} attribute is highly correlated with the
\textit{chaotic\_sleep} attribute, while the \textit{body\_mass\_indicator} 
attribute is negatively correlated with the \textit{cerebrovascular\_accident}
attribute.

The Cramér's V matrix in Figures \ref{fig:cramer_v_matrix_example_salary} and
\ref{fig:cramer_v_matrix_example_stroke} provides insights into the association
between discrete nominal attributes. For example, in the Salary Prediction dataset,
the \textit{gtype} attribute is strongly associated with the \textit{gender} attribute,
while in the Stroke Prediction dataset, there are no strong associations between
the discrete nominal attributes.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/correlation_matrix_SalaryPrediction_full.png}
    \caption{Correlation matrix for the Salary Prediction dataset}
    \label{fig:correlation_matrix_example_salary}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/correlation_matrix_AVC_full.png}
    \caption{Correlation matrix for the Stroke Prediction dataset}
    \label{fig:correlation_matrix_example_stroke}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/cramer_v_matrix_SalaryPrediction_full.png}
    \caption{Cramér's V matrix for the Salary Prediction dataset}
    \label{fig:cramer_v_matrix_example_salary}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/cramer_v_matrix_AVC_full.png}
    \caption{Cramér's V matrix for the Stroke Prediction dataset}
    \label{fig:cramer_v_matrix_example_stroke}
\end{figure}

@@

% ---- Bibliography ----
\begin{thebibliography}{8}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Correlation}{Wikipedia - Correlation}
    \bibitem{}
    \href{https://thinkingneuron.com/how-to-measure-the-correlation-between-two-categorical-variables-in-python/}{Thinking Neuron - How to Measure the Correlation Between Two Categorical Variables in Python}
    \bibitem{}
    \href{https://www.stratascratch.com/blog/chi-square-test-in-python-a-technical-guide/}{StrataScratch - Chi-Square Test in Python: A Technical Guide}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Cram\%C3\%A9r\%27s\_V}{Wikipedia - Cramér's V}
    \bibitem{}
